---
title: "Hands-on Exercise 1"
author: "Eugene Toh"
---

# Getting Started

## Installing and launching R packages

The code chunk below uses `p_load()` of `pacman` to check if `tidyverse` and `sf` are installed in the computer. If they are, they will be loaded by R.

```{r}
pacman::p_load(sf, tidyverse)
```

## Importing data

### Master Plan 2014 Subzone Boundary (Shapefile)

```{r}
mpsz = st_read(dsn = "data/geospatial/Master_Plan_2014_Subzone", layer = "MP14_SUBZONE_WEB_PL")
```
### Cycling Path 2014 (Shapefile)
```{r}
cyclingpath = st_read("data/geospatial/Cycling_Path_2014", layer = "CyclingPathGazette")
```
### Pre-school Locations (KML)
```{r}
preschool = st_read("data/geospatial/PreSchool_Location/PreSchoolsLocation.kml")
```
# Analysing the data
### Extract geometry of a simple features object
`mpsz` is a data frame-like like object where one of the columns contains geometric data. `st_geometry` returns a list of geometries. 
```{r}
st_geometry(mpsz)
```
### Extract attribute information of a simple features object
```{r}
glimpse(mpsz)
```
### Return the first few rows of a data frame
```{r}
head(mpsz, n=5)
```
### Getting the coordinate system of the feature object
```{r}
st_crs(mpsz)
```
## Adjusting the data
### Adjusting the EPSG code
```{r}
mpsz3414 <- st_set_crs(mpsz, 3414)
```
```{r}
st_crs(mpsz3414)
```
### Transforming the coordinate system
```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```
```{r}
st_crs(preschool3414)
```
## Plotting the data
```{r}
plot(mpsz)
```
### Plotting a single attribute
```{r}
plot(mpsz["PLN_AREA_N"])
```
## Working with Aspatial data
Aspatial data is data that is not geospatial in nature but contains fields that capture the x- and y-coordinates of the data points.

### Importing the data
```{r}
listings <- read_csv("data/aspatial/SG_Airbnb_Listing_Data/listings.csv")
```
### Reading the data
```{r}
list(listings)
```
#### Extracting the coordinates of each listing from the csv file and converting it into a data frame
```{r}
listings_sf <- st_as_sf(listings, coords = c("longitude", "latitude"), crs=4326) %>% st_transform(crs=3414)
```
Note that `%>%` is the pipe operator in R.
`EPSG: 4326` is `wgs84` and `EPSG: 3414` is Singapore SVY21 Projected Coordinate System. Refer to [the EPSG website](https://epsg.io) for details.

### Listing each row of the csv file
```{r}
glimpse(listings_sf)
```
# Geoprocessing
## Buffering
Getting the total area of land required for widening the existing cycling path by 5 metres:
```{r}
buffer_cycling <- st_buffer(cyclingpath, dist=5, nQuadSegs=30)
buffer_cycling$AREA <- st_area(buffer_cycling)
sum(buffer_cycling$AREA)
```
## Point-in-polygon count
Getting the number of pre-schools in each planning sub-zone.
`st_intersects` merge both the sub-zone and the pre-school data, while `lengths` does the counting.
```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
summary(mpsz3414$`PreSch Count`)
glimpse(mpsz3414)
```
Getting the districts with the most number of pre-schools.
```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```
Calculate the density of pre-schools of each district.
```{r}
mpsz3414$Area <- mpsz3414 %>% st_area()
mpsz3414 <- mpsz3414 %>% mutate(`PreSch Density` = `PreSch Count` / Area * 1000000)
glimpse(mpsz3414)
```
# Plotting with ggplot2
```{r}
hist(mpsz3414$`PreSch Density`)
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
ggplot(data=mpsz3414, 
       aes(y = `PreSch Count`, 
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black", 
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```