---
title: "Hands-on Exercise 7/8"
author: "Eugene Toh"
---

# Geographical Segmentation with Spatially Constrained Clustering Techniques

In geobusiness and spatial policy, it is common to divide a market into homogeneous regions using multivariate data. This exercise focuses on delineating Shan State, Myanmar, into regions based on ICT indicators: radio, television, landline phones, mobile phones, computers, and home internet.

## Installing packages

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

## Importing data

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

```{r}
summary(ict)
```

## Data wrangling

Since the townships with relatively higher total number of households tend to have higher number of households owning radio, TV, etc., we need to normalise the rate of ICT use to derive the penetration rate.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

## Exploratory data analysis

We can visualize the distribution of variables (e.g., the number of households with a radio) using appropriate Exploratory Data Analysis (EDA) techniques, as demonstrated in the code chunk below.

A histogram is useful for identifying the overall data distribution, such as left skew, right skew, or normal distribution.

This `ggplot` code creates a histogram to visualize the distribution of the `RADIO` variable from the `ict_derived` dataset.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

A right-skewed histogram (also known as positively skewed) means that the majority of the `RADIO` variable values are concentrated on the lower end (toward the left side of the histogram), with fewer observations on the higher end (toward the right). This indicates that most households in the dataset have relatively low values for radio ownership, while a smaller number of households have higher values.

In other words, there is a long tail on the right, showing that a few data points (households) have significantly higher values compared to the majority. This kind of skewness suggests that radio ownership is not evenly distributed across households, with many having low or moderate access and only a few having very high access.

Boxplot is useful to detect if there are outliers.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

If the boxes in the boxplot are positioned toward the left, it indicates that the majority of the values for the `RADIO` variable are relatively low. In a boxplot:

-   The central box represents the interquartile range (IQR), which contains the middle 50% of the data.

-   The line inside the box represents the median (the middle value).

-   The "whiskers" extend to the smallest and largest values within a certain range (typically 1.5 times the IQR from the box).

When the box is on the left side, it suggests that both the median and the majority of the data (the middle 50%) are concentrated at lower values for the `RADIO` variable. This implies that most households have relatively low radio ownership. If the whiskers are also short on the right side, it reinforces that higher values are less frequent, potentially indicating a right-skewed distribution.

Next, we will plot the distribution of the newly derived variables (e.g., radio penetration rate) using the code chunk below.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Next, the [*ggarrange()*](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html) function of [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/index.html) package is used to group these histograms together.

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

### EDA via choropleth map

Before we can prepare the choropleth map, we need to combine both the geospatial data object (i.e. *shan_sf*) and aspatial data.frame object (i.e. *ict_derived*) into one.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
qtm(shan_sf, "RADIO_PR")
```

To account for the bias in the distribution shown in the choropleth map, which is influenced by the total number of households in the townships, we will create two separate choropleth maps: one for the total number of households (`TT_HOUSEHOLDS.map`) and one for the total number of households with radios (`RADIO.map`). The code chunk below will be used to generate these maps.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

It is evident from the choropleth maps above that townships with a relatively larger number of households also tend to have higher radio ownership.

Let’s now plot the choropleth maps showing the distribution of the total number of households and the radio penetration rate using the code chunk below.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

Places with a lower total number of households tend to have a higher radio penetration rate.

## Correlation analysis

Before conducting cluster analysis, it is essential to ensure that the cluster variables are not highly correlated.

In this section, you will learn how to use the `corrplot.mixed()` function from the corrplot package to visualize and analyze the correlation between the input variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The correlation plot above indicates that COMPUTER_PR and INTERNET_PR are highly correlated, suggesting that only one of these variables should be used in the cluster analysis rather than both.

## Hierarchy cluster analysis

### Extracting clustering variables

The code chunk below will extract the clustering variables from the `shan_sf` simple feature object into a `data.frame`.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Note that the final list of clustering variables excludes `INTERNET_PR` due to its high correlation with `COMPUTER_PR`.

Next, we will modify the row labels to use township names instead of row numbers by applying the code chunk below.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Notice that the row number has been replaced into the township name.

Now, we will delete the TS.x field by using the code chunk below.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

### Data standardisation

In general, multiple variables will be used in cluster analysis. It is not unusual their values range are different. In order to avoid the cluster analysis result is baised to clustering variables with large values, it is useful to standardise the input variables before performing cluster analysis.

#### Min-max standardisation

In the code chunk below, *normalize()* of [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/index.html) package is used to stadardisation the clustering variables by using Min-Max method. The *summary()* is then used to display the summary statistics of the standardised clustering variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

Notice that the values range of the Min-max standardised clustering variables are 0-1 now.

#### Z-score standardisation

Z-score standardisation can be performed easily by using [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of Base R. The code chunk below will be used to stadardisation the clustering variables by using Z-score method.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

Notice the mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively.

**Note:** [*describe()*](https://www.rdocumentation.org/packages/Hmisc/versions/4.4-0/topics/describe) of [**psych**](https://cran.r-project.org/web/packages/psych/index.html) package is used here instead of *summary()* of Base R because the earlier provides standard deviation.

***Warning: Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.***

### Visualising the standardised clustering variables

Beside reviewing the summary statistics of the standardised clustering variables, it is also a good practice to visualise their distribution graphical.

The code chunk below plot the scaled *Radio_PR* field.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

-   **Comparison of Distribution Shapes**: The shape of the distribution in the raw data is preserved in the Min-Max standardized histogram, though it is scaled to a fixed range. The Z-score standardization, however, centers the data around zero, showing how many standard deviations each value deviates from the mean.

-   **Handling of Outliers**: The Z-score standardization may better highlight outliers because values far from the mean will show up more distinctly as they deviate by multiple standard deviations.

-   **Scaling for Analysis**: Min-Max standardization is useful when you need to keep the relative ranking of values intact (e.g., for algorithms that require bounded input), whereas Z-score standardization is helpful when analyzing data that assumes a normal distribution or when comparing across variables with different units.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### Computing proximity matrix

In R, many packages provide functions to calculate distance matrix. We will compute the proximity matrix by using [*dist()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of R.

*dist()* supports six distance proximity calculations, they are: **euclidean, maximum, manhattan, canberra, binary and minkowski**. The default is *euclidean* proximity matrix.

The code chunk below is used to compute the proximity matrix using *euclidean* method.

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

The code chunk below can then be used to list the content of *proxmat* for visual inspection.

```{r}
proxmat
```

### Computing hierarchical clustering

In R, there are several packages provide hierarchical clustering function. In this hands-on exercise, [*hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) of R stats will be used.

*hclust()* employed agglomeration method to compute the cluster. Eight clustering algorithms are supported, they are: ward.D, ward.D2, single, complete, average(UPGMA), mcquitty(WPGMA), median(WPGMC) and centroid(UPGMC).

The code chunk below performs hierarchical cluster analysis using ward.D method. The hierarchical clustering output is stored in an object of class **hclust** which describes the tree produced by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

We can then plot the tree by using *plot()* of R Graphics as shown in the code chunk below.

```{r}
plot(hclust_ward, cex = 0.6)
```

### Selecting the optimal clustering algorithm

One of the challenge in performing hierarchical clustering is to identify stronger clustering structures. The issue can be solved by using use [*agnes()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes) function of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like *hclus()*, however, with the *agnes()* function you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

The code chunk below will be used to compute the agglomerative coefficients of all hierarchical clustering algorithms.

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output above, we can see that Ward’s method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward’s method will be used.

### Determining optimal clusters

Another technical challenge face by data analyst in performing clustering analysis is to determine the optimal clusters to retain.

There are [three](https://statweb.stanford.edu/~gwalther/gap) commonly used methods to determine the optimal clusters, they are:

-   [Elbow Method](https://en.wikipedia.org/wiki/Elbow_method_(clustering))

-   [Average Silhouette Method](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)

-   [Gap Statistic Method](http://www.web.stanford.edu/~hastie/Papers/gap.pdf)

#### Gap statistic method

The [**gap statistic**](http://www.web.stanford.edu/~hastie/Papers/gap.pdf) compares the total within intra-cluster variation for different values of k with their expected values under null reference distribution of the data. The estimate of the optimal clusters will be value that maximize the gap statistic (i.e., that yields the largest gap statistic). This means that the clustering structure is far away from the random uniform distribution of points.

To compute the gap statistic, [*clusGap()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/clusGap) of [**cluster**](https://cran.r-project.org/web/packages/cluster/index.html) package will be used.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Also note that the [*hcut*](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

Next, we can visualise the plot by using [*fviz_gap_stat()*](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

```{r}
fviz_gap_stat(gap_stat)
```

With reference to the gap statistic graph above, the recommended number of cluster to retain is 1. However, it is not logical to retain only one cluster. By examine the gap statistic graph, the 6-cluster gives the largest gap statistic and should be the next best cluster to pick.

**Note:** In addition to these commonly used approaches, the [NbClust](https://cran.r-project.org/web/packages/NbClust/) package, published by Charrad et al., 2014, provides 30 indices for determining the relevant number of clusters and proposes to users the best clustering scheme from the different results obtained by varying all combinations of number of clusters, distance measures, and clustering methods.

### Interpreting the dendrograms

In the dendrogram displayed above, each leaf corresponds to one observation. As we move up the tree, observations that are similar to each other are combined into branches, which are themselves fused at a higher height.

The height of the fusion, provided on the vertical axis, indicates the (dis)similarity between two observations. The higher the height of the fusion, the less similar the observations are. Note that, conclusions about the proximity of two observations can be drawn only based on the height where branches containing those two observations first are fused. We cannot use the proximity of two observations along the horizontal axis as a criteria of their similarity.

It’s also possible to draw the dendrogram with a border around the selected clusters by using [*rect.hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/rect.hclust.html) of R stats. The argument *border* is used to specify the border colors for the rectangles.

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### Visually-driven hierarchical clustering analysis

In this section, we will learn how to perform visually-driven hiearchical clustering analysis by using [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/index.html) package.

With **heatmaply**, we are able to build both highly interactive cluster heatmap or static cluster heatmap.

#### Transforming the data frame into a matrix

The data was loaded into a data frame, but it has to be a data matrix to make your heatmap.

The code chunk below will be used to transform *shan_ict* data frame into a data matrix.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### Plotting interactive cluster heatmap using \`heatmaply\`

In the code chunk below, the [*heatmaply()*](https://talgalili.github.io/heatmaply/reference/heatmaply.html) of [heatmaply](https://talgalili.github.io/heatmaply/index.html) package is used to build an interactive cluster heatmap.

```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### Mapping the clusters formed

With closed examination of the dendragram above, we have decided to retain six clusters.

[*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The output is called *groups*. It is a *list* object.

In order to visualise the clusters, the *groups* object need to be appended onto *shan_sf* simple feature object.

The code chunk below form the join in three steps:

-   the *groups* list object will be converted into a matrix;

-   *cbind()* is used to append *groups* matrix onto shan_sf to produce an output simple feature object called `shan_sf_cluster`; and

-   *rename* of **dplyr** package is used to rename *as.matrix.groups* field as *CLUSTER*.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

Next, *qtm()* of **tmap** package is used to plot the choropleth map showing the cluster formed.

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

The choropleth map above reveals the clusters are very fragmented. The is one of the major limitation when non-spatial clustering algorithm such as hierarchical cluster analysis method is used.

# Spatially Constrained Clustering

## SKATER approach

### Converting into `SpatialPolygonsDataFrame`

First, we need to convert `shan_sf` into a `SpatialPolygonsDataFrame` because the `SKATER` function only supports `sp` objects, such as `SpatialPolygonsDataFrame`. The code snippet below uses the `as_Spatial()` function from the `sf` package to convert `shan_sf` into a `SpatialPolygonsDataFrame` named `shan_sp`.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### Computing neighbour list

Next, the `poly2nb()` function from the `spdep` package will be used to generate the neighbors list from the polygon data.

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

Now we need to plot the boundaries and visualise the nearest neighbour relationships.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

Since we are using `plot`, the individual map layers and how they are drawn is determined by which comes first.

### Computing minimum spanning tree

#### Calculating edge costs

Next, the `nbcosts()` function from the `spdep` package is used to calculate the cost of each edge, which represents the distance between its nodes. This function computes the distance using a data frame with observation vectors for each node.

The code snippet below is used to calculate the cost of each edge.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

For each observation, this provides the pairwise dissimilarity between its values for the five variables and those of its neighboring observations (based on the neighbor list). Essentially, this represents a generalized weight for a spatial weights matrix.

Next, we will incorporate these costs into a weights object, similar to how we calculated inverse distance weights. Specifically, we convert the neighbor list into a list weights object by assigning the previously computed `lcosts` as the weights.

To achieve this, the `nb2listw()` function from the `spdep` package is used, as demonstrated in the code snippet below.

Note that we specify the style as "B" to ensure the cost values are not row-standardized.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

### Computing minimum spanning tree

The minimum spanning tree is computed by mean of the [*mstree()*](https://r-spatial.github.io/spdep/reference/mstree.html) of **spdep** package as shown in the code chunk below.

```{r}
shan.mst <- mstree(shan.w)
```

After computing the MST, we can check its class and dimension by using the code chunk below.

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

Note that the dimension is 54 and not 55. This is because the minimum spanning tree consists on n-1 edges (links) in order to traverse all the nodes.

We can display the content of *shan.mst* by using *head()* as shown in the code chunk below.

```{r}
head(shan.mst)
```

The plot method for the MST include a way to show the observation numbers of the nodes in addition to the edge. As before, we plot this together with the township boundaries. We can see how the initial neighbour list is simplified to just one edge connecting each of the
nodes, while passing through all the nodes.

```{r}
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

### Computing spatially constrained clusters using SKATER method

The code snippet below calculates the spatially constrained clusters using the `skater()` function from the `spdep` package.

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

The `skater()` function requires three mandatory arguments:

-   the first two columns of the MST matrix (excluding the cost),

-   the data matrix (used to update costs as units are grouped),

-   and the number of cuts, which is set to one less than the desired number of clusters (i.e., the specified value represents the number of cuts in the graph, not the number of clusters).

The output of `skater()` is an object of class `skater`. You can inspect its contents using the code snippet below.

```{r}
str(clust6)
```

The most interesting component of this list structure is the groups vector containing the labels of the cluster to which each observation belongs (as before, the label itself is arbitary). This is followed by a detailed summary for each of the clusters in the edges.groups list. Sum of squares measures are given as ssto for the total and ssw to show the effect of each of the cuts on the overall criterion.

We can check the cluster assignment by using the conde chunk below.

```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observations are in each cluster by means of the table command. Parenthetially, we can also find this as the dimension of each vector in the lists contained in edges.groups. For example, the first list has node with dimension 12, which is also the number of observations in the first cluster.

```{r}
table(ccs6)
```

Lastly, we can also plot the pruned tree that shows the five clusters on top of the townshop area.

```{r}
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

### Visualising the clusters in choropleth map

The following code snippet is used to visualize the clusters generated by the SKATER method.

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

To make comparison simpler, it's more effective to display the hierarchical clustering map alongside the spatially constrained hierarchical clustering map.

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5)
shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5)
tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

## ClustGeo approach

In this section, you'll get practical experience with the ClustGeo package, using its functions to conduct both traditional hierarchical clustering and spatially constrained cluster analysis.

The ClustGeo package in R is designed for spatially constrained cluster analysis, offering a Ward-like hierarchical algorithm called `hclustgeo()`, which incorporates spatial/geographical constraints. The algorithm uses two dissimilarity matrices, D0 (for attribute space) and D1 (for constraint space), and a mixing parameter alpha (ranging from 0 to 1). D0 can be non-Euclidean, and observation weights can be non-uniform. The goal is to find an alpha value that increases spatial contiguity while maintaining solution quality, assisted by the `choicealpha()` function.

The ClustGeo package offers the `hclustgeo()` function for Ward-like hierarchical clustering, similar to `hclust()`. For non-spatially constrained clustering, simply provide a dissimilarity matrix, as shown in the code snippet below.

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

To perform spatially constrained hierarchical clustering, a spatial distance matrix is derived using `st_distance()` from the `sf` package, with `as.dist()` converting the data frame to a matrix.

Next, `choicealpha()` determines a suitable alpha value. Based on the graphs above, we’ll use alpha = 0.2. Then, `cutree()` generates the cluster object, which is joined back to the `shan_sf` polygon feature data frame.

Finally, the newly defined spatially constrained clusters can be plotted.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```

## Visual interpretation of clusters

The code below shows the distribution of the clustering variable `RADIO_PR` by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

The boxplot indicates that Cluster 3 has the highest mean radio ownership per thousand households, followed by Clusters 2, 1, 4, 6, and 5.

Studies have shown that parallel coordinate plots effectively reveal clustering variables by cluster. In the code below, `ggparcoord()` from the GGally package is used.

```{r}
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

The plot shows that households in Cluster 4 own the most TVs and mobile phones, while Cluster 5 households own the least of all five ICTs.

The `scale` argument in `ggparcoord()` offers several scaling methods:

-   `std`: standardizes by subtracting the mean and dividing by the standard deviation.

-   `robust`: uses the median and median absolute deviation.

-   `uniminmax`: scales variables between 0 and 1.

-   `globalminmax`: no scaling; range is set by the global min and max.

-   `center`: centers each variable after scaling.

-   `centerObs`: centers each variable at a specific observation.

There’s no single best method; choose the one that fits your analysis.

Lastly, summary statistics (mean, median, etc.) can complement the visual interpretation. In the code below, `group_by()` and `summarise()` from `dplyr` calculate the mean values for the clustering variables.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```
