---
title: "In-class Exercise 2"
author: "Eugene Toh"
date: "last-modified"
execute:
  freeze: true
---

```{r}
pacman::p_load(tidyverse, sf, tmap)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial")
```

-   Simple features are represented via "simple features geometry (SFG)", and they and geospatial data can be represented with "simple features objects (SFO)" which represent a collection of SFG as a data-frame object. The non-spatial data (metadata) are stored in a separate column separate from the SFG column. The data type of that SFG column is known as "simple features collection (SFC)".

-   SFO is an object, you can get the class by passing in a variable to the `class` function.

-   In the environments tab, "obs" stands for observations which is basically the number of rows. "Variables" represent the number of columns.

-   The data types you will be working with for each column would be:

    -   `int`

    -   `chr`

        -   Basically a string.

    -   `Date`

    -   `num`

        -   A superclass of `int` and `double`. R will automatically cast it to its child classes if needed, but this is basically a generic number class.

```{r}
mpsz2019 <- st_read(dsn = "data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% st_transform(crs = 3414)
```

If you look at the KML version, you can see that it is less tidier than the Shapefile version.

```         
<center><table><tr><th colspan='2' align='center'><em>Attributes</…
```

You'll have to manually parse and extract the data. Hence, the Shapefile version will be the one we are going to use.

```{r}
tm_shape(mpsz)+
  tm_fill("REGION_N", 
          style = "quantile",
          palette = "plasma",
          title = "Subzones") +
  tm_layout(main.title = "Planning subzones of Singapore (2014)",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)")
```

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.kml")
preschool3414 <- st_transform(preschool, crs = 3414)
```

```{r}
popdata <- st_read("data/aspatial/respopagesextod2023.csv")
popdata2023 <- popdata %>%
  group_by(PA, SZ, AG)
```
