---
title: "In-class Exercise 4"
---

```{r}
pacman::p_load(sf, raster, spatstat, sparr, tmap, tidyverse)
```

```{r}
geography <- st_read("data/rawdata") %>% st_union() %>% st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 32748)
```

Polygon Z: polygons with height data, you need to remove it.

Converting it into owin:

```{r}
kbb_owin <- as.owin(geography)
kbb_owin
```

Checking whether it is really an owin:

```{r}
class(kbb_owin)
```

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>% st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% st_transform(crs = 32748) # do not switch order of latlong
```

Now, you want separate fields for day, month, and year for easier data filtering. The `DayOfYear` field takes the number of days in a year into account.

```{r}
fire_sf <- fire_sf %>% mutate(DayOfYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, label = TRUE, abbr = FALSE))
```

```{r}
tm_shape(geography) +
  tm_polygons() +
  tm_shape(fire_sf) +
  tm_dots()
```

Use `tm_symbols` or `tm_markers` if you have a custom point logo.

Polygon \> line \> point.

A facet is a set of maps.

```{r}
tm_shape(geography) +
  tm_polygons() +
  tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
  tm_facets(by = "Month_fac", free.coords = FALSE, drop.units = TRUE)
```

`free.coords` make sure that each map has the same map extent.

Extracting by month:

```{r}
fire_month <- fire_sf %>% select(Month_num)
```

Creating ppp (to use `spatstat`):

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
summary(fire_month_ppp)
```

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

Spatio-temporal KDE:

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

Most of the fires started from July onwards:

```{r}
#| fig-width: 12
#| fig-height: 10 # use this if things are too cramped for graphs
tims <- c(7, 8, 9, 10, 11, 12)
par(mfcol=c(2, 3))
for (i in tims) {
  plot(st_kde, i, override.par=FALSE, fix.range=TRUE, main = paste("KDE at month", i))
}
```

Instead of using month, you can use day which can make the value of lambda more meaningful.
